---
title: Reproducable Research - John Hopkins Data Scientist Certification - Program
  assignment 1
author: "Ron Collins"
date: "Sunday, January 18,2015"
output: pdf_document
---
> what is this


1. Loading libraries necessary for R
```{r}
require("knitr")
require("ggplot2")
require("plyr")
require("dplyr")
require("RCurl")
require("grid")
require("gridExtra")
```

2. Input data for anslysis
    * Create temp file
    * assign URL to download from
    * Download zip file into temp file
    * unzip downloaded file
    * unlink temp file

```{r}
echo = TRUE
temp <- tempfile()
URL <- "http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip" 
download.file(URL,temp)
proj1_unzip <- unzip(temp)
unlink(temp)

```
3. Read zip file
```{r}
echo = TRUE
PG1_data <- read.table(proj1_unzip, header = TRUE, sep = ",")
```


For this part of the assignment, you can ignore the missing values in the dataset.

4. Make a histogram of the total number of steps taken each day
    *Clean up dataframe by removing missing values
```{r}
#clean up data set for missing values
na_vector <- !is.na(PG1_data$steps)
PG1_data_clean <- PG1_data[na_vector,]
```
    * calculating the total number of steps for each day
    
```{r}

# Change PG1_data_clean$date to factor for analysis
PG1_data_clean$date <- as.factor(PG1_data_clean$date)

# Caculate the total steps taken for each date
total_steps <- summarise(group_by(PG1_data_clean, date),sum=sum(steps))

# Convert PG1_data_clean$date from factor back to date
PG1_data_clean$date <- as.Date(PG1_data_clean$date)

```

    * plotting bar plot using ggplot2 for total number of steps each day
    
```{r}

t_s <- ggplot(data=total_steps, aes(x=date, y=sum, fill=time)) + 
    geom_bar(colour="black", fill="black", width=.7, stat="identity") + 
    guides(fill=FALSE) +
    xlab("Date") + ylab("Total Number of Steps") +
    ggtitle("Total steps per day")

t_s +  theme(axis.title.x = element_text(face="bold", colour="black", size=12),
            axis.text.x  = element_text(angle=90, vjust=0.5, size=10)) 


```

   5. Calculate and report the mean and median number of steps for each day
    
```{r}
# Change PG1_data_clean$interval to factor for analysis
PG1_data_clean$date <- as.factor(PG1_data_clean$date)
avg_median_steps_day <- summarise(group_by(PG1_data_clean, date), avg =mean(steps), med = median(steps))
avg_median_steps_day$date <- as.Date(avg_median_steps_day$date)
``` 
    

   6. What is the average daily activity pattern?
        * Open graphics device for windows
        * ggplot bar plot average number of steps per day
        * scale_x_date(format = "%b-%Y") +
        scale_x_date(breaks =NULL,labels = date_format("%d-%b")) +
    
```{r}
library(scales) # load scales

t_s <- ggplot(data=avg_median_steps_day, aes(x=date, y=avg, fill=time)) + 
    geom_bar(colour="black", fill="black", width=.7, stat="identity") +      
    guides(fill=FALSE) + 
    xlab("Date") + ylab("Average Number of Steps") + 
    ggtitle("Average number of steps per day")

t_s +  theme(axis.title.x = element_text(face="bold", colour="black", size=12),
            axis.text.x  = element_text(angle=90, vjust=0.5, size=10)) 

```

    7. Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, 
       averaged across all days (y-axis)
       * calculate average number of steps taken per 5 minute interval
       * make time series plot
       
        
```{r}

# convert PG1_data_clean$interval to factor for summarising
PG1_data_clean$interval <- as.factor(PG1_data_clean$interval)


# calculate avg number of steps per sampling interval using summarise function    
avg_median_steps_interval<- summarise(group_by(PG1_data_clean, interval), avg =mean(steps), med = median(steps))
avg_median_steps_interval$interval <- as.numeric(avg_median_steps_interval$interval)

# make time series plott using ggplot2

ggplot(avg_median_steps_interval, aes(interval, avg)) + geom_line() +
  xlab("Intervals") + ylab("Average Steps") +  ggtitle("Time series plot of the average number of steps over 5 minute intervals")


```

   8.  Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
        *
```{r}
max_no_steps <-    max(avg_median_steps_interval$avg)
for (i in 1 : length(avg_median_steps_interval$avg)) {   
    if(avg_median_steps_interval[i,2] == max_no_steps) max_avg_step_interval <- i
}
output_txt <- paste0(" The maximum average step occured during the ", max_avg_step_interval, "the time interval.")
print(output_txt)
 
```

Imputing missing values

Note that there are a number of days/intervals where there are missing values (coded as NA). The presence of missing days may introduce bias into some calculations or summaries of the data.

9. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)
    
```{r}
na_vector <- is.na(PG1_data$steps)
na_number <- sum(na_vector)
na_txt <- paste0(" The number of missing observations in the dataset with steps per interval on days is ", na_number, ".")
print(na_txt)

```

10. Correct for missing data
    * Fill in all of the missing values in the dataset by calculating the mean for the 5-minute intervals and
        substutinting for the missing values.
        
    * Create a new dataset that is equal to the original dataset but with the missing data filled in.
        
```{r}
PG1_data$interval <- as.factor(PG1_data$interval)
avg_steps <- summarise(group_by(PG1_data, interval), avg =mean(steps, na.rm=TRUE))
interval_avg <- data.frame(interval=integer(288), avg = integer(288))
interval_avg$interval <- avg_steps$interval
interval_avg$avg <- avg_steps$avg
interval_avg$avg <- as.integer(avg_steps$avg)
# convert interval_avg$interval from factor to numeric
interval_avg$interval <- as.numeric(interval_avg$interval)



PG1_data_comp <- PG1_data

# convert date and interval from factor to date and numberic
PG1_data_comp$date <- as.Date(PG1_data_comp$date)
PG1_data_comp$interval <- as.numeric(PG1_data_comp$interval)

# Substitute calcualted average value for missing step values
for(j in 1:length(PG1_data_comp$date)) {
    if(is.na(PG1_data_comp[j,1])) {
        raw_interval <- PG1_data_comp[j,3]
        for(k in 1:288){
         if(raw_interval == interval_avg[k,1])  {
             PG1_data_comp[j,1] <- interval_avg[k,2]
         } 
        }
    }
}
na_vector <- is.na(PG1_data_comp$steps)
na_number <- sum(na_vector)
na_txt <- paste0(" The number of observations in the dataset with NA step value, after manipulation of data set, per interval on days is ", na_number, ".")
print(na_txt)

```
   
11. Calculate and report the mean and median total number of  steps taken per day for the corrected data.  
    * What is the impact of imputing missing data on the estimates of the total daily number of steps?
    

        
```{r}
# calcualate the total, mean and median for data set with missing values replaced
total_steps_corr <- summarise(group_by(PG1_data_comp, date),sum=sum(steps), avg =mean(steps), med = median(steps))
# print out dataframe of results

```

12. Do these values differ from the estimates from the first part of the assignment? 
* Cleaned dataframe is total_steps.   dataframe with average replacing missing value is total_steps_corr.
```{r}
print("The value of the estimates differ from the first assignment when comparing the number of dates.") 
print (" There are 53 dates present in the cleaned dataframe and 61 dates presented in the corrected dataframs.")


# ttest to compare means of steps taken
t.test(total_steps$sum, total_steps_corr$sum) 

print("There are no differences between the means of the corrected and cleaned dataframe")

```


13. Make a histogram of the total number of steps taken each day and . 
```{r}
# Produce bar plot for total steps taken each day during study

t_s <- ggplot(data=total_steps_corr, aes(x=date, y=sum, fill=time)) + 
    geom_bar(colour="black", fill="black", width=.7, stat="identity") + 
    guides(fill=FALSE) +
    xlab("Date") + ylab("Total Number of Steps") +
    ggtitle("Total steps per day")

t_s +  theme(axis.title.x = element_text(face="bold", colour="black", size=12),
            axis.text.x  = element_text(angle=90, vjust=0.5, size=10)) 


```



14. Are there differences in activity patterns between weekdays and weekends?

    *For this part the weekdays() function may be of some help here. Use the dataset with the filled-in missing values 
        for this part.

   * Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given 
    date is a weekday or weekend day.
    ```{r}
# Create new dataframe with day of week and weekday or weekend.
day_of_week <- total_steps_corr
weekday <- total_steps_corr$date
dayType <- total_steps_corr$date
day_of_week <- cbind(day_of_week,weekday)
day_of_week <- cbind(day_of_week,dayType)
day_of_week$dayType <- as.character(day_of_week$dayType)
#convert factors class to native class
day_of_week$date <- as.Date(day_of_week$date)
day_of_week$weekday <- as.character(day_of_week$weekday)
day_of_week$dayType <- as.integer(day_of_week$dayType)
week_day_types <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
week_end_types <- c("Saturday", "Sunday")

# Loop to convert date to Weedday or weekend
for (k in 1:length(day_of_week$date)){
    day_of_week[k,5] <- weekdays(day_of_week[k,1])
    for (j in 1:5){
       if(day_of_week[k,5]== week_day_types[j]) day_of_week[k,6]<-"Weekday"
    }
    for(l in 1:2){
        if(day_of_week[k,5]==week_end_types[l])day_of_week[k,6] <- "weekend"
    }
}

# Use general linear model to compare steps between weekend and weekday steps
# comvert dayType from character to factor for analysis
# Analysis for average number of steps
day_of_week$dayType <-as.factor(day_of_week$dayType)
fit_avg_steps <- glm(day_of_week$avg~day_of_week$dayType)
summary(fit_avg_steps)

# Analysis for total steps
fit_total_steps <- glm(day_of_week$sum~day_of_week$dayType)
summary(fit_total_steps)
```

    The number of average and total number of steps for weekend and weekday are not statistically different at a 0.05 and 0.01 level.  The weekend and weekday are statistically different at the 0.1 level.
    

15. Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average
    number of steps taken, averaged across all weekday days or weekend days (y-axis). The plot should look something 
    like the following, which was creating using simulated data:

    *Your plot will look different from the one above because you will be using the activity monitor data. Note that
    the above plot was made using the lattice system but you can make the same version of the plot using any plotting 
    system you choose.
    ```{r}V  
# time series of average steps taken for each time series interval
plot_1 <- ggplot(avg_median_steps_interval, aes(interval, avg)) + geom_line() +
        xlab("Intervals") + ylab("Average Steps") +  ggtitle("Time series plot of the average number of  
        steps over 5 minute intervals")
plot_1

# Plot average number of steps across all weedend days(y- axis)
# Calculate the average number of steps by weekday and weekend.
day_of_week$dayType <- as.factor(day_of_week$dayType)
average_steps_dayType <- summarise(group_by(day_of_week, dayType),avg=mean(avg))
# convert dayType from factor to character
average_steps_dayType$dayType <-as.character(average_steps_dayType$dayType)

            
plot_2 <- qplot(factor(dayType), data=average_steps_dayType, geom="bar", fill=factor(avg)) 

plot_2 <- ggplot(data=average_steps_dayType, aes(x=dayType, y=avg)) + geom_bar(colour="black", stat="identity") + ggtitle("Average number of steps for Weekday and weekend") + xlab("Day of Week") + ylab("Average Steps") 
        
plot_2

# Plot both plots on one graph
library(grid)
library(gridExtra)
grid.arrange(plot_1, plot_2, ncol=1, main = "Two graphs Combined onto one page" )
```
    
